---
title: "Getting Started with datalab"
author: "lut hasan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with datalab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(datalab)
```

# Introduction

Welcome to `datalab`! This package helps you explore and visualize data without worrying about complex coding. You can create beautiful charts and compute statistics with simple, easy-to-remember functions.

## Quick Start

### Step 1: Load Your Data

First, load your data and call it `data`:

```{r}
# Load the sample agriculture dataset
data(agriculture_data)
data <- agriculture_data
```

### Step 2: Explore Your Variables

See what's in your data:

```{r}
head(data)
```

### Step 3: Create Your First Chart

Make a line chart to see the relationship between two variables:

```{r}
linechart(nitrogen_content, crop_yield)
```

## More Examples

### Boxplots

Show the distribution of a single variable:

```{r}
# Load heart disease data
data(heart_data)
data <- heart_data

boxplot(age)
```

Compare groups:

```{r}
boxplot(age, group = 'sex')
```

### Pie Charts

Show how categories are distributed:

```{r}
piechart(chest_pain_type)
```

Split by another variable to compare:

```{r}
piechart(chest_pain_type, 'sex')
```

### Descriptive Statistics

Get a summary of your variables:

```{r}
descriptives(age, resting_bp, cholesterol)
```

### Frequency Tables

Count how many times each value appears:

```{r}
frequencies(chest_pain_type)
```

## Tips for Workshop Participants

### Using Variables

You can assign variables to letters to make typing easier:

```{r eval=FALSE}
x = nitrogen_content
y = crop_yield
linechart(x, y)
```

### Changing Variables

Want to look at different variables? Just reassign:

```{r eval=FALSE}
a = chest_pain_type
b = age
piechart(a, b)
```

### Working with Your Own Data

To use your own data:

1. Load your data file (CSV or Excel)
2. Assign it to `data`
3. Use the functions as shown above

```{r eval=FALSE}
# For CSV files
data <- read.csv("myfile.csv")

# For Excel files (requires readxl package)
library(readxl)
data <- read_excel("myfile.xlsx")
```

## Common Problems and Solutions

### "Variable not found" error

**Problem:** You see an error like `Variable 'age' not found in data`

**Solution:** Check that:
- You spelled the variable name correctly
- You loaded the data and called it `data`
- The variable exists in your dataset

### "No data frame found" error

**Problem:** You see `No data frame found`

**Solution:** Make sure you loaded your data and assigned it to `data`:

```{r eval=FALSE}
data <- my_dataset
```

### Charts look cluttered

**Problem:** Too many categories in a pie chart

**Solution:** Use a frequency table first to see your data:

```{r}
frequencies(chest_pain_type)
```

## One-Slide Demo Script (for Facilitators)

**Goal:** Show participants how to go from data to insight in 5 minutes

1. **Load data** (30 seconds)
   ```{r eval=FALSE}
   data(heart_data)
   data <- heart_data
   ```

2. **Look at data** (30 seconds)
   ```{r eval=FALSE}
   head(data)
   ```

3. **Make a chart** (1 minute)
   ```{r eval=FALSE}
   piechart(chest_pain_type)
   ```

4. **Compare groups** (1 minute)
   ```{r eval=FALSE}
   piechart(chest_pain_type, sex)
   ```

5. **Get statistics** (1 minute)
   ```{r eval=FALSE}
   descriptives(age, resting_bp)
   ```

6. **Discuss insight** (1.5 minutes): "What story does this data tell?"

## Next Steps

Try these exercises on your own:

**Exercise 1:** Create a boxplot comparing cholesterol levels between males and females.

**Exercise 2:** Make a frequency table of chest pain types, sorted from most to least common.

**Exercise 3:** Create a line chart showing the relationship between age and resting blood pressure.

---

## Reference: All Available Functions

- `linechart(x, y)` - Line chart
- `boxplot(var)` - Boxplot  
- `piechart(var, filter)` - Pie chart(s)
- `descriptives(...)` - Summary statistics
- `frequencies(var)` - Frequency table

For help with any function, type `?function_name` (e.g., `?linechart`)